Xolotl: An Intuitive and Approachable Neuron and Network Simulator for Research and Teaching

### 2018-08-17 paper/paper.tex
minor typos in manuscript

	Conductance-based models of neurons are used extensively in computational neuroscience, and working with these models continues to be a challenge due to their high dimensionality and large number of parameters. Here, we present a neuron and network simulator built on a novel automatic type system that binds object-oriented code written in C++ to objects in MATLAB. Our approach builds on the tradition of uniting the speed of languages like C++ with the ease-of-use and feature-set of scientific programming languages like MATLAB. This software allows for the creation and manipulation of hierarchical models with components that are named and searchable, permitting intuitive high-level programmatic control over all parameters. The simulator's architecture allows for the interactive manipulation of any parameter in any model, and for visualizing the effects of changing that parameter immediately. xolotl is fully featured with hundreds of ion channel models from the electrophysiological literature, and can be extended to include arbitrary conductances, synapses, and mechanisms. Several core features like bookmarking of parameters and automatic hashing of source code facilitate reproducible and auditable research. Its ease of use and rich visualization capabilities make it an attractive option in teaching environments. Finally, xolotl is written in a modular fashion, includes detailed tutorials and worked examples, and is freely available at https://github.com/sg-s/xolotl, enabling seamless integration into the workflow of other researchers.

@@

To overcome these design limitations, we have developed a novel automatic type system, that we call cpplab, which binds MATLAB code to classes specified in C++ header files. This architecture automatically creates objects in MATLAB that represent the underlying object-oriented C++ code, allowing the symbolic manipulation of C++ objects in the MATLAB interface. In this paper, we introduce xolotl, an implementation of the cpplab system specialized in integrating conductance-based neuron and network models. Models can be easily constructed from network building blocks in a few lines of MATLAB code using a hierarchical and intuitive syntax. Since models in the MATLAB workspace are automatically linked to models in the C++ implementation, configuring these objects in MATLAB transparently configures the underlying C++ objects. xolotl comes packaged with hundreds of components that can be used to assemble cells and networks;  has built-in visualization functions to inspect voltage time traces and activation functions; and a GUI for real-time manipulation of model parameters. The ease-of-use of the tools built into our software lends them to pedagogical applications and the rapid prototyping of models. Our software aims to simplify the investigation of the dynamics of conductance-based network and neuron models, facilitate collaborative modeling, and is intended to complement other tools being developed in the computational neuroscience community.

@@

xolotl is designed to be easy-to-use and richly featured while being fast enough to use in everyday research. Our focus was on designing an approachable simulator of conductance-based neurons and networks of these neurons; simulating arbitrary dynamical systems is therefore beyond the scope of this software. Specifically, the software was designed to simulate models of the form

@@

This software is designed to be used from within MATLAB, a scientific programming language common amongst neuroscientists and engineers for pedagogy and research. Our goal was to make xolotl completely usable entirely from within MATLAB. Models created using this simulator appear in the MATLAB workspace as native objects, are are thus fully scriptable, and are fully compatible with the large library of toolboxes that MATLAB provides, allowing the software to be used as a component of other packages and tools. All parameters of models and activation functions of any channel can be inspected at any point. We designed several features of the simulator to be easily extensible: adding custom conductances or synapse types is possible by calling functions that generate new C++ files on-the-fly. Finally, xolotl fully auditable by design, with several tools to verify model and parameter integrity and aid in reproducibility.

@@

\paragraph{Object-oriented.}
xolotl is designed to mirror the nested and hierarchical structure of networks and neurons. Biological neuronal networks are made up of neurons that are connected to each other with synapses. Each of these neurons contains within it a set of conductances and synaptic currents that contribute to its electrical behavior. Intracellular mechanisms can act within the cell, or parts of the cell, to modify and regulate conductances, synapses, or other dynamic properties of the cell. Similarly, a xolotl model can contain a set of compartments that can represent individual neurons. Each compartment can contain an arbitrary set of conductances. Compartments, conductances and synapses can contain mechanisms that can affect anything in the model. All objects in the xolotl model tree (compartments, conductances, etc.) are bonafide MATLAB objects with their own type and properties. This object-oriented programming paradigm naturally represents the hierarchical structure of biological networks that contain neurons that contain populations of ion channels, and makes constructing, integrating, and thinking about these models easier.

\paragraph{A rich library of network components.} xolotl comes packaged with hundreds of pre-existing components (compartments, synapses, conductances, and mechanisms) that can be used as building blocks to construct model neurons and networks. Compartments represent sections of membrane with a single membrane potential, Calcium level, and set of constituent components, representing either entire cells or parts of cells. Objects of type "conductance" represent populations of ion channels in a compartment that produce transmembrane currents. "synapse" objects connect two compartments together by introducing a current in the postsynaptic compartment that depends on the presynaptic compartment's membrane potential. Objects of type "mechanism" can represent any intracellular mechanism and can read and modify any other component in the cell, and can run arbitrary dynamical systems within them.

@@

\paragraph{Automatic hashing and compiling.}
Because every model requires a compiled binary to run, a potential stumbling block is the problem of unambiguously linking a model to a binary executable. xolotl solves this problem by hashing (\cite{rivestMD5MessageDigestAlgorithm1992}) the C++ header files of every component in the model recursively, allowing a model, no matter how complex, to be compactly represented by a short alphanumeric identifier (its ``hash''). Compiled binaries are named using this hash, ensuring both that the correct binary is run to integrate the model, and that compilation occurs only as needed. This powerful feature enables the user experience to remain entirely within the MATLAB workspace, with compilation and selection of the correct binary occurring silently in the background.

@@

\paragraph{Limited numerical integration strategies.} Most components in the software are integrated using the exponential Euler method, which has been shown to perform well in neuronal models (\cite{ohErrorAnalysisSpecialized2006, dayanTheoreticalNeuroscience2001}). However, it may be desirable to use other methods under certain conditions. It is possible to introduce new network components that implement other integration schemes, or to modify the integration schemes of existing components, but that requires writing new C++  code. Currently, xolotl can only implement integration schemes with fixed step size.

\paragraph{Inefficient tools for handling large networks.} xolotl was designed to work with small but complex networks and models, where every compartment and component is named, rather than numbered. It is more therefore suited towards simulating small and heterogeneous networks rather than large but homogenous networks. While the software can integrate large networks ($>1000$ compartments), other tools are presumably more suited to this task, offering more natural frameworks for dealing with a large number of identical units. Similarly, xolotl is not optimized to solve coupled ODEs on complex branched morphologies, and other simulators like NEURON (\cite{hinesNEURONSimulationEnvironment1997}) are more suited for this task.

@@

In this example, we demonstrate how to simulate the spiking activity of a Hodgkin-Huxley-like model, and how the tools built into xolotl make it easy to set up and integrate the model and gain insight into the underlying biophysical mechanisms. This simple model consists of a single electrical compartment with three types of conductances (Figure \ref{fig:figurehh}A). This hierarchical organization of the neuron is mirrored in the structure of the model in the simulator: an object of type "compartment" is used to represent the cell body, and three objects of type "conductance" are used to model the three populations of ion channels (Figure \ref{fig:figurehh}B). These models of ion channels were obtained from (\cite{liuModelNeuronActivityDependent1998}) based on electrophysiological recordings from the lobster stomatogastric ganglion (\cite{turrigianoSelectiveRegulationCurrent1995}), and are part of the simulator. The code to set up this model is terse, idiomatic, relies on no special markup, and preserves the hierarchical nature of the model (Figure \ref{fig:figurehh}C).

@@

The model can be integrated for various amplitudes of injected current to determine its F-I (frequency current) curve, or the relationship between injected current and its firing rate (\cite{kisperskyIncreaseSodiumConductance2012}). Figure \ref{fig:figurehh}E shows the f-I curve of this model, obtained by repeated integration of the model.  Finally, the built-in \texttt{show} function can plot activation ($m$) and inactivation ($h$) curves and voltage-dependent timescales of any channel type in the simulator (Figure \ref{fig:figurehh}D-G). These plots reveal that activation kinetics of the "NaV" channels are much faster than that of the "Kd" channels (Figure \ref{fig:figurehh}F), which facilitates the transient depolarization in an action potential. In summary, the simulator allows the user to construct and integrate this model in a few lines of code, and provides rich visualization of the dynamics of the model.

@@

Voltage clamping is an experimental technique where a amplifier is configured to inject the appropriate amount of current through a electrode to maintain the voltage of a cell at a desired value (\cite{dayanTheoreticalNeuroscience2001}). Under this paradigm, the membrane voltage is "clamped" or fixed to a desired value, permitting the study of voltage-dependent ion channels, since the sum of all currents through the population of ion channels in the cell is equal and opposite to the current injected by the clamp (Figure \ref{fig:figureclamp}A). By combining voltage clamp with the use of pharmacological agents to block all channels but the one of interest, the voltage-sensitivity of an ion channel population can be characterized (\cite{coleIonicCurrentMeasurements1960, coleIonsPotentialsNerve1955, hodgkinEffectSodiumIons1949, hodgkinMeasurementCurrentvoltageRelations1952, hodgkinQuantitativeDescriptionMembrane1952, turrigianoSelectiveRegulationCurrent1995}).

xolotl can reproduce such a voltage clamp experiment \textit{in-silico}. Figure \ref{fig:figureclamp}B illustrates how a simple model with a single compartment and a single ion channel type can be set up and clamped to a desired voltage.  Integrating the model yields the current required to clamp the cell at that voltage. Here, we use a delayed-rectifier potassium conductance (\cite{liuModelNeuronActivityDependent1998}) and simulate a voltage-clamp experiment whose goal is to infer the activation function of this channel. First, the cell is clamped to a number of different voltages (Fig. \ref{fig:figureclamp}C) and the resultant clamp currents are measured by integrating the model (Fig. \ref{fig:figureclamp}D). Since the compartment is being voltage clamped, integrating the model returns the clamping current:

@@

By repeating the integration at a number of clamp voltages, we observe that the asymptotic clamp currents depend on the clamp voltage in a nonlinear manner (Fig. \ref{fig:figureclamp}E), since the open probabilities of the channel are functions of the membrane voltage. Assuming the reversal potential is known, Eq. \eqref{eq:2} can be used to solve for the total conductance of the channel as a function of the clamp voltage (Fig. \ref{fig:figureclamp}F). Finally, a sigmoid can be fit to the normalized conductance-voltage curves to obtain the activation function of the ion channel population (Fig. \ref{fig:figureclamp}G-H). xolotl can therefore be used to describe graphically the theoretical underpinnings of ion channel characterization through voltage clamp and can serve as an effective pedagogical tool in computational neuroscience.

@@

So far, the models we described only considered the voltage dynamics of a cell (the solution to Eq. \ref{eq:1}). However, real neurons possess several dynamical features, arising from a variety of intracellular mechanisms. xolotl makes it possible to model and include arbitrary intracellular mechanisms. In xolotl, these intracellular mechanisms are represented by the "mechanism" object, and can be bound to compartments, conductances, and other object types.

@@

Switching back and forth points in parameter space and state space of a neuron model is a common occurrence in working with neuron models, and a significant fraction of a modeler's time is spent in a feedback loop of running simulations, viewing the output, changing parameters, and re-running simulations (\cite{deschutterConsumerGuideNeuronal1992}). xolotl makes it easy to bookmark configurations of a model and return to them at will. Internally, xolotl uses the "serialize" method to gather all parameters and dynamic variables into a vector of values that is passed to the underlying C++ implementation. A paired "deserialize" method is used to update all parameters and variables in the object tree from this vector. This architecture provides a natural framework for representing the state of any model, no matter how complex, using a vector of numbers. The "snapshot" method built into xolotl leverages this schema to save the entire state of the model in a named variable, that can be accessed using another built-in method called "reset".

@@

xolotl has several types of synapses like Cholinergic, Glutamatergic, AMPA and NMDA built-in, and other synapse classes can be easily added using templates. Figure \ref{fig:figurenetwork} demonstrates the implementation of a model of the triphasic pyloric rhythm in the stomatogastric ganglion of crustaceans (\cite{prinzSimilarNetworkActivity2004}). The pyloric model contains three compartments ("AB", "LP", and "PY") and seven synapses (Figure \ref{fig:figurenetwork}A). This structure is recapitulated in the hierarchy of the xolotl object, where conductances are contained within compartments (Figure \ref{fig:figurenetwork}B). The membrane potentials show triphasic rhythmicity in the three compartments (Figure \ref{fig:figurenetwork}C-E). When the "PY" is depolarized, the dynamical variable mediating the  glutamatergic ("Glut") synapse model between "PY" and "LP" is close to 1 and "LP" is inhibited (Figure \ref{fig:figurenetwork}F). Conversely, when "PY" is hyperpolarized, the dynamical variable is close to 0. When "PY" spikes, IPSPs can be seen in the "LP" voltage trace (Figure \ref{fig:figurenetwork}D-E).

@@

Conductance-based neuron models are typically high dimensional and contain many parameters. Changing a single parameter monotonically can cause non-monotonic changes in behavior of the model, and certain dynamical features may only emerge when in specific non-convex regions of parameter space (\cite{golowasch2002failure}). It is often challenging to build intuition about what effect a parameter has on the model under these conditions. Traditionally, the technique used by computational neuroscientists in building intuition about these models is to iteratively run simulations, view outputs and change parameters. In practice, this meant writing a script, running it, inspecting the output, changing parameters in the script, and repeating this process. It can be cumbersome, and every step in this process involves ``mode'' changes: between a text editor, viewing a graphical output, and the command line that can frustrate the researcher.

@@

This feature is only possible due to our architectural decision to split the code base across two programming environments. A rich scientific programming environment like MATLAB makes it possible to generate user interface elements and to bind them to data in plots, while the sheer speed of compiled languages like C++ allow for the immediate, real-time feedback and updating of plots. By default, any parameter in the model can be manipulated, including parameters in user-defined mechanisms that do not exist in the base simulator. The GUI can be constrained to arbitrary subsets of model parameters using wild card matching (as shown in Fig. \ref{fig:figuremanipulate}) or by manually specifying the parameters of interest.

@@

Many simulators have been designed with a focus on simulate large numbers of compartments, either as networks with many identical neurons or in a large multi-compartment neuron model (\cite{bretteSimulationNetworksSpiking2007, sherfeyDynaSimMATLABToolbox2018, vitayANNarchyCodeGeneration2015, delormeSpikeNETEventdrivenSimulation2003}). While our software is not designed for this task {\em per se}, we measured its performance as a function of the number of compartments simulated. xolotl can quickly create a number of identical copies of a compartment using the "replicate" method:

@@

We used the "replicate" method to create and run models with varying numbers of neurons (either Hodgkin-Huxley-like or bursting neurons) and measured the speed of all three simulators as a function of system size. Plotting the speed normalized by the system size vs. the system size, we observed that the speed of integration of xolotl is linear with system size (Figure \ref{fig:figurebenchmark}G-H, black lines), for up to 1000 single-compartment neurons (up to 13,000 ODEs). Its performance compares favorably with that of NEURON and DynaSim as a function of system size (Figure \ref{fig:figurebenchmark}G-H).

@@

We set out to design a neuron and network simulator that could be useful in the classroom setting, especially for students of computational science, while also being powerful, fast, and extensible enough to be used for research. By using a novel architecture that permits the symbolic manipulation of C++ objects in a intuitive MATLAB interface, we demonstrated some of the features of xolotl using simulations of single-compartment models of Hodgkin-Huxley like neurons (Fig. \ref{fig:figurehh}); voltage clamp experiments to recover activation functions of single channel types (Fig. \ref{fig:figureclamp}); a neuron model where intracellular mechanisms can control the dynamics of Calcium and can regulate the maximal conductance of ion channel types in an activity-dependent manner (Fig. (\ref{fig:figuremechanism}); and a network of neurons with multiple synapse types (Fig. \ref{fig:figurenetwork}). We also illustrated how built-in features of the simulator make it easy to bookmark and jump between model configurations (Fig. \ref{fig:figuremechanism}, purple), and how parameters of the model can be changed using sliders and their effect can be viewed in real time (Fig. \ref{fig:figuremanipulate}).

@@

``About half the time spent on a typical simulation project involves creating and tuning the model. Thus, a good user interface may contribute more to the overall efficiency of a project than pure computation speed.'' (\cite{deschutterConsumerGuideNeuronal1992}). xolotl is designed primarily with ease-of-use in mind. This includes the time it takes to install, setup, and learn how to use the software, the time to write and debug scripts, and the time to perform the simulations (\cite{rudolph2007much}). An easy-to-use simulation environment must minimize time spent in all these domains, especially during human engagement with the software. Complicated software remains broadly inaccessible and time-consuming even to perform single-compartment simulations, though the actual simulation time may be very small.

@@

Over the years, several simulators have been developed to integrate systems of coupled differential equations, those that model the spiking activity of neurons (\cite{bretteSimulationNetworksSpiking2007, sherfeyDynaSimMATLABToolbox2018, vitayANNarchyCodeGeneration2015, delormeSpikeNETEventdrivenSimulation2003, hinesNEURONSimulationEnvironment1997, bower2003genesis}). A critical architectural choice in designing a simulator is how much ``scaffolding'' a user is provided to construct a model; whether a model is specified by equations or by components, or by some combination of the two. In an equation-oriented architecture, the user starts with a blank slate and the primary method of specifying a model is to write out its differential equations. In contrast, in a component-oriented architecture, the primary method of specifying a model is to assemble it from pre-existing components, each of which include differential equations, parameters, and solvers. Both approaches have advantages and disadvantages that are illustrated below.

@@

Another architectural choice in simulators is the syntax required for specifying models. Equation-oriented simulators specify models as strings of equations, so must invent a new syntax to specify derivatives, variables, and other common elements in coupled differential equations. Some component-oriented simulators like NEURON also specify their own syntax, or have invented their own language to specify components and write out equations. While this allows for powerful features such as support for units in NEURON, the user is required to learn a novel syntax and vocabulary, hindering ease of use. In general, the syntax for model specification in different simulators can be different. For example, DynaSim, BRIAN, XPP, and NEURON all use different, incompatible formalisms to represent equations, increasing the cognitive load on users using more than one simulator. Here, we have elected not to specify a domain-specific ``middleware'' layer, and instead specify and implement models and equations in idiomatic C++. This greatly decreases the learning curve and allows users with a general familiarity with programming languages to quickly acquaint themselves even with the most technical parts of the simulator.

>>>>

	Conductance-based models of neurons are used extensively in computational neuroscience, and working with these models continues to be a challenge due to their high dimensionality and large number of parameters. Here, we present a neuron and network simulator built on a novel automatic type system that binds object-oriented code written in C++ to objects in MATLAB. Our approach builds on the tradition of uniting the speed of languages like C++ with the ease-of-use and feature-set of scientific programming languages like MATLAB. Xolotl allows for the creation and manipulation of hierarchical models with components that are named and searchable, permitting intuitive high-level programmatic control over all parts of the model. The simulator's architecture allows for the interactive manipulation of any parameter in any model, and for visualizing the effects of changing that parameter immediately. Xolotl is fully featured with hundreds of ion channel models from the electrophysiological literature, and can be extended to include arbitrary conductances, synapses, and mechanisms. Several core features like bookmarking of parameters and automatic hashing of source code facilitate reproducible and auditable research. Its ease of use and rich visualization capabilities make it an attractive option in teaching environments. Finally, xolotl is written in a modular fashion, includes detailed tutorials and worked examples, and is freely available at https://github.com/sg-s/xolotl, enabling seamless integration into the workflow of other researchers.

@@

To overcome these design limitations, we have developed a novel automatic type system, that we call cpplab, which binds MATLAB code to classes specified in C++ header files. This architecture automatically creates objects in MATLAB that represent the underlying object-oriented C++ code, allowing the symbolic manipulation of C++ objects in the MATLAB interface. In this paper, we introduce xolotl, an implementation of the cpplab system specialized in integrating conductance-based neuron and network models. Models can be easily constructed from components of different types in a few lines of MATLAB code using a hierarchical and intuitive syntax. Since models in the MATLAB workspace are automatically linked to models in the C++ implementation, configuring these objects in MATLAB transparently configures the underlying C++ objects. Xolotl comes packaged with hundreds of components that can be used to assemble cells and networks;  has built-in visualization functions to inspect voltage time traces and activation functions; and a GUI for real-time manipulation of model parameters. Xolotl's ease of use makes it an attractive option for pedagogical applications, rapid prototyping of models, and primary research use. Our software aims to simplify the investigation of the dynamics of conductance-based network and neuron models, facilitate collaborative modeling, and is intended to complement other tools being developed in the computational neuroscience community.

@@

Xolotl is designed to be easy-to-use and richly featured while being fast enough to use in everyday research. Our focus was on designing an approachable simulator of conductance-based neurons and networks of these neurons; simulating arbitrary dynamical systems is therefore beyond the scope of this software. Specifically, the software was designed to simulate models of the form

@@

This software is designed to be used from within MATLAB, a scientific programming language common amongst neuroscientists and engineers for pedagogy and research. Our goal was to make xolotl completely usable entirely from within MATLAB. Models created using this simulator appear in the MATLAB workspace as native objects, are thus fully scriptable, and are fully compatible with the large library of toolboxes that MATLAB provides, allowing the software to be used as a component of other packages and tools. All parameters of a model, and activation functions of any channel can be inspected at any point. We designed several features of the simulator to be easily extensible: adding custom conductances or synapse types is possible by calling functions that generate new C++ files on-the-fly. Finally, xolotl is fully auditable by design, with several tools to verify model and parameter integrity and aid in reproducibility.

@@

\paragraph{Object-oriented.}
Xolotl is designed to mirror the nested and hierarchical structure of networks and neurons. Biological neuronal networks are made up of neurons that are connected to each other with synapses. Each of these neurons contains within it a set of conductances and synaptic currents that contribute to its electrical behavior. Intracellular mechanisms can act within the cell, or parts of the cell, to modify and regulate conductances, synapses, or other dynamic properties of the cell. Similarly, a xolotl model can contain a set of compartments that can represent individual neurons. Each compartment can contain an arbitrary set of conductances. Compartments, conductances and synapses can contain mechanisms that can affect anything in the model. All objects in the xolotl model tree (compartments, conductances, etc.) are bonafide MATLAB objects with their own type and properties. This object-oriented programming paradigm naturally represents the hierarchical structure of biological networks that contain neurons that contain populations of ion channels, and makes constructing, integrating, and thinking about these models easier.

\paragraph{A rich library of network components.} Xolotl comes packaged with hundreds of pre-existing components (compartments, synapses, conductances, and mechanisms) that can be used as building blocks to construct model neurons and networks. ``Compartments'' represent sections of membrane with a single membrane potential, intracellular Calcium concentration, and set of constituent components; and can represent either entire cells or parts of cells. Objects of type ``conductance'' represent populations of ion channels in a compartment that produce transmembrane currents. ``synapse'' objects connect two compartments together by introducing a current in the postsynaptic compartment that depends on the presynaptic compartment's membrane potential. Objects of type ``mechanism'' can represent any intracellular mechanism and can read and modify any other component in the cell, and can run arbitrary dynamical systems within them.

@@

\paragraph{Automatic hashing and compiling.}
Because every model requires a compiled binary to run, a potential stumbling block is the problem of unambiguously linking a model to a binary executable. Xolotl solves this problem by hashing (\cite{rivestMD5MessageDigestAlgorithm1992}) the C++ header files of every component in the model recursively, allowing a model, no matter how complex, to be compactly represented by a short alphanumeric identifier (its ``hash''). Compiled binaries are named using this hash, ensuring both that the correct binary is run to integrate the model, and that compilation occurs only as needed. This powerful feature enables the user experience to remain entirely within the MATLAB workspace, with compilation and selection of the correct binary occurring silently in the background.

@@

\paragraph{Limited numerical integration strategies.} Most components in the software are integrated using the exponential Euler method, which has been used in integrating neuronal models (\cite{ohErrorAnalysisSpecialized2006, dayanTheoreticalNeuroscience2001}). However, it may be desirable to use other methods under certain conditions. It is possible to introduce new components that implement other integration schemes, or to modify the integration schemes of existing components, but that requires writing new C++  code. Currently, xolotl can only implement integration schemes with fixed step size.

\paragraph{Inefficient tools for handling large networks.} xolotl was designed to work with small but complex networks and models, where every compartment and component is named, rather than numbered. It is more therefore suited towards simulating small, heterogeneous networks rather than large, homogenous networks. While the software can integrate large networks ($>1000$ compartments), other tools are presumably more suited to this task, offering more natural frameworks for dealing with a large number of identical units. Similarly, xolotl is not optimized to solve coupled ODEs on complex branched morphologies, that other simulators like NEURON (\cite{hinesNEURONSimulationEnvironment1997}) are specialized for.

@@

In this example, we demonstrate how to simulate the spiking activity of a Hodgkin-Huxley-like model, and how the tools built into xolotl make it easy to set up and integrate the model and gain insight into the underlying biophysical mechanisms. This simple model consists of a single electrical compartment with three types of conductances (Figure \ref{fig:figurehh}A). This hierarchical organization of the neuron is mirrored in the structure of the model in the simulator: an object of type ``compartment'' is used to represent the cell body, and three objects of type ``conductance'' are used to model the three populations of ion channels (Figure \ref{fig:figurehh}B). These models of ion channels were obtained from (\cite{liuModelNeuronActivityDependent1998}) based on electrophysiological recordings from the lobster stomatogastric ganglion (\cite{turrigianoSelectiveRegulationCurrent1995}), and are part of the simulator. The code to set up this model is terse, idiomatic, relies on no special markup, and preserves the hierarchical nature of the model (Figure \ref{fig:figurehh}C).

@@

The model can be integrated for various amplitudes of injected current to determine its F-I (frequency current) curve (\cite{kisperskyIncreaseSodiumConductance2012}). Figure \ref{fig:figurehh}E shows the F-I curve of this model, obtained by repeated integration of the model.  Finally, the built-in \texttt{show} function can plot activation ($m$) and inactivation ($h$) curves and voltage-dependent timescales of any channel type in the simulator (Figure \ref{fig:figurehh}D-G). These plots reveal that activation kinetics of the "NaV" channels are much faster than that of the "Kd" channels (Figure \ref{fig:figurehh}F), which facilitates the transient depolarization in an action potential. In summary, the simulator allows the user to construct and integrate this model in a few lines of code, and provides rich visualization of the dynamics of the model.

@@

Voltage clamping is an experimental technique where a amplifier is configured to inject the appropriate amount of current through a electrode to maintain the voltage of a cell at a desired value (\cite{dayanTheoreticalNeuroscience2001}). Under this paradigm, the membrane voltage is ``clamped'' or fixed to a desired value, permitting the study of voltage-dependent ion channels, since the sum of all currents through the population of ion channels in the cell is equal and opposite to the current injected by the clamp (Figure \ref{fig:figureclamp}A). By combining voltage clamp with the use of pharmacological agents to block all channels but the one of interest, the voltage-sensitivity of an ion channel population can be characterized (\cite{coleIonicCurrentMeasurements1960, coleIonsPotentialsNerve1955, hodgkinEffectSodiumIons1949, hodgkinMeasurementCurrentvoltageRelations1952, hodgkinQuantitativeDescriptionMembrane1952, turrigianoSelectiveRegulationCurrent1995}).

Xolotl can reproduce such a voltage clamp experiment \textit{in-silico}. Figure \ref{fig:figureclamp}B illustrates how a simple model with a single compartment and a single ion channel type can be set up and clamped to a desired voltage.  Integrating the model yields the current required to clamp the cell at that voltage. Here, we use a delayed-rectifier potassium conductance (\cite{liuModelNeuronActivityDependent1998}) and simulate a voltage-clamp experiment whose goal is to infer the activation function of this channel. First, the cell is clamped to a number of different voltages (Fig. \ref{fig:figureclamp}C) and the resultant clamp currents are measured by integrating the model (Fig. \ref{fig:figureclamp}D). Since the compartment is being voltage clamped, integrating the model returns the clamping current:

@@

By repeating the integration at a number of clamp voltages, we observe that the asymptotic clamp currents depend on the clamp voltage in a nonlinear manner (Fig. \ref{fig:figureclamp}E), since the open probabilities of the channel are functions of the membrane voltage. Assuming the reversal potential is known, Eq. \eqref{eq:2} can be used to solve for the total conductance of the channel as a function of the clamp voltage (Fig. \ref{fig:figureclamp}F). Finally, a sigmoid can be fit to the normalized conductance-voltage curves to obtain the activation function of the ion channel population (Fig. \ref{fig:figureclamp}G-H). Xolotl can therefore be used to describe graphically the theoretical underpinnings of ion channel characterization through voltage clamp and can serve as an effective pedagogical tool in computational neuroscience.

@@

So far, the models we described only considered the voltage dynamics of a cell (the solution to Eq. \ref{eq:1}). However, real neurons possess several dynamical features, arising from a variety of intracellular mechanisms. Xolotl makes it possible to model and include arbitrary intracellular mechanisms. In xolotl, these intracellular mechanisms are represented by the ``mechanism'' object, and can be bound to compartments, conductances, and other object types.

@@

Switching back and forth points in parameter space and state space of a neuron model is a common occurrence in working with neuron models, and a significant fraction of a modeler's time is spent in a feedback loop of running simulations, viewing the output, changing parameters, and re-running simulations (\cite{deschutterConsumerGuideNeuronal1992}). Xolotl makes it easy to bookmark configurations of a model and return to them at will. Internally, xolotl uses the "serialize" method to gather all parameters and dynamic variables into a vector of values that is passed to the underlying C++ implementation. A paired "deserialize" method is used to update all parameters and variables in the object tree from this vector. This architecture provides a natural framework for representing the state of any model, no matter how complex, using a vector of numbers. The "snapshot" method built into xolotl leverages this schema to save the entire state of the model in a named variable, that can be accessed using another built-in method called "reset".

@@

Xolotl has several types of synapses built-in, and other synapse classes can be easily added using templates. Figure \ref{fig:figurenetwork} demonstrates the implementation of a model of the triphasic pyloric rhythm in the stomatogastric ganglion of crustaceans (\cite{prinzSimilarNetworkActivity2004}). The pyloric model contains three compartments ("AB", "LP", and "PY") and seven synapses (Figure \ref{fig:figurenetwork}A). This structure is recapitulated in the hierarchy of the xolotl object, where conductances are contained within compartments (Figure \ref{fig:figurenetwork}B). The membrane potentials show triphasic rhythmicity in the three compartments (Figure \ref{fig:figurenetwork}C-E). When the "PY" is depolarized, the dynamical variable mediating the  glutamatergic ("Glut") synapse model between "PY" and "LP" is close to 1 and "LP" is inhibited (Figure \ref{fig:figurenetwork}F). Conversely, when "PY" is hyperpolarized, the dynamical variable is close to 0. In this model, when "PY" spikes, IPSPs can be seen in the "LP" voltage trace (Figure \ref{fig:figurenetwork}D-E).

@@

Conductance-based neuron models are typically high dimensional and contain many parameters. Changing a single parameter monotonically can cause non-monotonic changes in behavior of the model, and certain dynamical features may only emerge when in specific non-convex regions of parameter space (\cite{golowasch2002failure}). It is often challenging to build intuition about what effect a parameter has on the model under these conditions. Traditionally, the technique used by computational neuroscientists in building intuition about these models is to iteratively run simulations, view outputs and change parameters. In practice, this meant writing a script, running it, inspecting the output, changing parameters in the script, and repeating this process. It can be cumbersome, and every step in this process involves ``mode'' changes: switching between a text editor, viewing a graphical output, and the command line that can frustrate the researcher.

@@

This feature was only possible due to our architectural decision to split the code base across two programming environments. A rich scientific programming environment like MATLAB makes it possible to easily generate user interface elements and to bind them to data in plots, while the sheer speed of compiled languages like C++ allow for the immediate, real-time feedback and updating of plots. By default, any parameter in the model can be manipulated, including parameters in user-defined mechanisms that do not exist in the base simulator. The GUI can be constrained to arbitrary subsets of model parameters using wild card matching (as shown in Fig. \ref{fig:figuremanipulate}) or by manually specifying the parameters of interest.

@@

Many simulators have been designed with a focus on simulate large numbers of compartments, either as networks with many identical neurons or in a large multi-compartment neuron model (\cite{bretteSimulationNetworksSpiking2007, sherfeyDynaSimMATLABToolbox2018, vitayANNarchyCodeGeneration2015, delormeSpikeNETEventdrivenSimulation2003}). While our software is not designed for this task {\em per se}, we measured its performance as a function of the number of compartments simulated. Xolotl can quickly create a number of identical copies of a compartment using the "replicate" method:

@@

We used the "replicate" method to create and run models with varying numbers of neurons (either Hodgkin-Huxley-like or bursting neurons) and measured the speed of all three simulators as a function of system size. Plotting the speed normalized by the system size {\em vs.} the system size, we observed that the speed of integration of xolotl is linear with system size (Figure \ref{fig:figurebenchmark}G-H, black lines), for up to 1000 single-compartment neurons (up to 13,000 ODEs). Its performance compares favorably with that of NEURON and DynaSim as a function of system size (Figure \ref{fig:figurebenchmark}G-H).

@@

We set out to design a neuron and network simulator that could be useful in the classroom setting, especially for students of computational neuroscience, while also being powerful, fast, and extensible enough to be used for research. By using a novel architecture that permits the symbolic manipulation of C++ objects in a intuitive MATLAB interface, we demonstrated some of the features of xolotl using simulations of single-compartment models of Hodgkin-Huxley like neurons (Fig. \ref{fig:figurehh}); voltage clamp experiments to recover activation functions of single channel types (Fig. \ref{fig:figureclamp}); a neuron model where intracellular mechanisms can control the dynamics of Calcium and can regulate the maximal conductance of ion channel types in an activity-dependent manner (Fig. (\ref{fig:figuremechanism}); and a network of neurons with multiple synapse types (Fig. \ref{fig:figurenetwork}). We also illustrated how built-in features of the simulator make it easy to bookmark and jump between model configurations (Fig. \ref{fig:figuremechanism}, purple), and how parameters of the model can be changed using sliders and their effect can be viewed in real time (Fig. \ref{fig:figuremanipulate}).

@@

``About half the time spent on a typical simulation project involves creating and tuning the model. Thus, a good user interface may contribute more to the overall efficiency of a project than pure computation speed.'' (\cite{deschutterConsumerGuideNeuronal1992}). Xolotl is designed primarily with ease-of-use in mind. This includes the time it takes to install, setup, and learn how to use the software, the time to write and debug scripts, and the time to perform the simulations (\cite{rudolph2007much}). An easy-to-use simulation environment must minimize time spent in all these domains, especially during human engagement with the software. Complicated software remains broadly inaccessible and time-consuming even to perform single-compartment simulations, though the actual simulation time may be very small.

@@

Over the years, several simulators have been developed to integrate systems of coupled differential equations that model the spiking activity of neurons (\cite{bretteSimulationNetworksSpiking2007, sherfeyDynaSimMATLABToolbox2018, vitayANNarchyCodeGeneration2015, delormeSpikeNETEventdrivenSimulation2003, hinesNEURONSimulationEnvironment1997, bower2003genesis}). A critical architectural choice in designing a simulator is how much ``scaffolding'' a user is provided with to construct a model; whether a model is specified by equations or by components, or by some combination of the two. In an equation-oriented architecture, the user starts with a blank slate and the primary method of specifying a model is to write out its differential equations (\cite{stimberg2014equation}). In contrast, in a component-oriented architecture, the primary method of specifying a model is to assemble it from pre-existing components, each of which include differential equations, parameters, and solvers. Both approaches have advantages and disadvantages that are discussed below.

@@

Another architectural choice in designing simulators is the syntax required for specifying models. Equation-oriented simulators specify models as strings of equations, so must invent a new syntax to specify derivatives, variables, and other common elements in coupled differential equations. Some component-oriented simulators like NEURON also specify their own syntax, or have invented their own language to specify components and write out equations. While this allows for powerful features such as support for units in NEURON, the user is required to learn a novel syntax and vocabulary, hindering ease of use. In general, the syntax for model specification in different simulators can be different. For example, DynaSim, BRIAN, XPP, and NEURON all use different, incompatible formalisms to represent equations, increasing the cognitive load on users using more than one simulator. Here, we have elected not to specify a domain-specific ``middleware'' layer, and instead specify and implement models and equations in idiomatic C++. This greatly decreases the learning curve and allows users with a general familiarity with programming languages to quickly acquaint themselves even with the most technical parts of the simulator.
